Phylogenetics and the human microbiome.

### 2013-10-07 sbreview.tex
Expanded OTU section H/T @b4winckler. Closes #4.
Björn-- in my reading, KS10 was about clustering "true" metagenomic
reads (rather than surveys) and so didn't quite fit. LMK if I missed
something. Thanks for the suggestions!

There are many OTU inference methods with various speeds and strategies.
Of the fixed-cutoff methods, the traditional choice has been CD-HIT \citep{li2006cdhit}, which seems to have been supplanted by clustering features of USEARCH \citep{edgar2010usearch} and perhaps now its descendant UPARSE \citep{edgar2013uparse}.
These methods are very fast and are heuristic in that they are described as an algorithm rather than as global optimization of some notion of goodness of clustering.
\citet{white2010alignment} show that different ways of doing this heuristic clustering can result in very different results.
\citet{navlakha2009finding} develop methods that try to come up with sequence groupings that more closely reflect the type of species divisions found in taxonomies.
\citet{hao2011clustering} use a Gaussian mixture model formulation for clustering to avoid fixed cutoff values.

\newcommand{\pdLegend}{
    Unweighted phylogenetic diversity (PD, left) and an abundance-weighted PD measure (right), where taxa present in a sample are shown as blue circles and abundances are shown as the size of the circles.
    Unweighted PD takes the sum of the branch lengths connecting samples.
    Abundance-weighted measures take a weighted sum of branch lengths where weight is determined in some way by the abundance of the taxa on either side of the branch: if we give edges width according to their weight, the abundance-weighted measure can be thought of as the sum of the areas of the edges.
}
\forarxiv{
\begin{wrapfigure}[23]{r}{2.5in}
  % \capstart
  \vspace{-27pt}
  \begin{center}
    \includegraphics[width=2.5in]{figures/pd.pdf}
  \end{center}
  \vspace{-10pt}
  \caption{\pdLegend}
  \label{fig:pd}
\end{wrapfigure}
}

@@

\newcommand{\unifracLegend}{
    The UniFrac divergence measure \citep[figure adapted from][]{LozuponeKnightUniFrac05}.
    When samples are interspersed across the tree (left tree), they have a smaller \textit{fraction} of branch lengths that can be \textit{uniquely} assigned to one sample or another compared to when they are separate (right tree).
}
\forarxiv{
\begin{wrapfigure}[23]{r}{2.5in}
  % \capstart
  \vspace{-16pt}
  \begin{center}
    \includegraphics[width=2.5in]{figures/unifrac.pdf}
  \end{center}
  \vspace{-10pt}
  \caption{\unifracLegend}
  \label{fig:unifrac}
\end{wrapfigure}
}

@@

@@

\newcommand{\dirtpilesLegend}{
    Part of a minimal mass movement to calculate the earth-mover's distance on a phylogenetic tree.
}
\forarxiv{
\begin{wrapfigure}[19]{l}{1.6in}
  % \capstart
% \vspace{-24pt}
  \begin{center}
  \end{center}
  \vspace{-5pt}
  \caption{\dirtpilesLegend}
  \label{fig:dirtpiles}
\end{wrapfigure}
}

There have been efforts to augment these ordination visualizations with additional information giving more structure to the visualizations.
Biplots display variables (in the microbial case summarized by taxonomic labels) as points along with points representing samples \citep[e.g.][]{hewitt2013bacterial,lozupone2013meta}.
\citet{PurdomAnalyzingDataGraphs08} describes how generalized principal component eigenvectors can be interpreted via weightings on the leaves of a phylogenetic tree.
\citet{matsen2013edge} have developed a variant of principal components analysis that explicitly labels the axes with weightings on phylogenetic trees that indicate their influence.

>>>>

There are many OTU inference methods with various speeds and strategies.
Of the fixed-cutoff methods, the traditional choice has been CD-HIT \citep{li2006cdhit}, which seems to have been supplanted by clustering features of USEARCH \citep{edgar2010usearch} and perhaps now its descendant UPARSE \citep{edgar2013uparse}.
These methods are very fast and are heuristic in that they are described as an algorithm rather than as global optimization of some notion of goodness of clustering.
\citet{white2010alignment} show that different ways of doing this heuristic clustering can result in very different results.
\citet{navlakha2009finding} and \citet{WYS13} develop methods that try to come up with sequence groupings that more closely reflect the type of species divisions found in taxonomies,
\citet{hao2011clustering} use a Gaussian mixture model formulation for clustering to avoid fixed cutoff values,
and \cite{CWC12} use a two step process, first with a Dirichlet multinomial mixtures on 3-mer profiles, and then a minimum description length criterion.

@@

\newcommand{\pdLegend}{
    Unweighted phylogenetic diversity (PD, left) and an abundance-weighted PD measure (right), where taxa present in a sample are shown as blue circles and abundances are shown as the size of the circles.
    Unweighted PD takes the sum of the branch lengths connecting samples.
    Abundance-weighted measures take a weighted sum of branch lengths where weight is determined in some way by the abundance of the taxa on either side of the branch: if we give edges width according to their weight, the abundance-weighted measure can be thought of as the sum of the areas of the edges.
}
\forarxiv{
\begin{wrapfigure}[25]{r}{2.2in}
  % \capstart
  \vspace{-14pt}
  \begin{center}
    \includegraphics[width=2.2in]{figures/pd.pdf}
  \end{center}
  \vspace{-10pt}
  \caption{\pdLegend}
  \label{fig:pd}
\end{wrapfigure}
}

@@

\newcommand{\unifracLegend}{
    The UniFrac divergence measure \citep[figure adapted from][]{LozuponeKnightUniFrac05}.
    When samples are interspersed across the tree (left tree), they have a smaller \textit{fraction} of branch lengths that can be \textit{uniquely} assigned to one sample or another compared to when they are separate (right tree).
}
\forarxiv{
\begin{wrapfigure}[23]{r}{2.2in}
  % \capstart
  \vspace{-26pt}
  \begin{center}
    \includegraphics[width=2.2in]{figures/unifrac.pdf}
  \end{center}
  \vspace{-10pt}
  \caption{\unifracLegend}
  \label{fig:unifrac}
\end{wrapfigure}
}

@@

\newcommand{\dirtpilesLegend}{
    Part of a minimal mass movement to calculate the earth-mover's distance on a phylogenetic tree.
}
\forarxiv{
\begin{wrapfigure}[14]{l}{1.2in}
  % \capstart
  \vspace{-17pt}
  \begin{center}
    \includegraphics[width=1.2in]{figures/tree_dirtpiles.pdf}
  \end{center}
  \vspace{-5pt}
  \caption{\dirtpilesLegend}
  \label{fig:dirtpiles}
\end{wrapfigure}
}

There have been several efforts to augment these ordination visualizations with additional information giving more structure to the visualizations.
Biplots display variables (in the microbial case summarized by taxonomic labels) as points along with points representing samples \citep[e.g.][]{hewitt2013bacterial,lozupone2013meta}.
\citet{PurdomAnalyzingDataGraphs08} describes how generalized principal component eigenvectors can be interpreted via weightings on the leaves of a phylogenetic tree.
\citet{matsen2013edge} have developed a variant of principal components analysis that explicitly labels the axes with weightings on phylogenetic trees that indicate their influence.
